---
title: "How to Install rstan and brms"
date: "`r Sys.Date()`"
---

## For Mac Users

* Make sure R is up-to-date > 4.2
* [Configural C++ Toolchain](https://github.com/stan-dev/rstan/wiki/Configuring-C---Toolchain-for-Mac)
  * Set up the C++ Toolchain using the [macOS R toolchain installer](https://github.com/coatless-mac/r-macos-rtools#how-do-i-use-the-installer)
  * To improve estimation speed, you can copy the following syntax to the R console
  
```{r}
#| eval: false
dotR <- file.path(Sys.getenv("HOME"), ".R")
if (!file.exists(dotR)) dir.create(dotR)
M <- file.path(dotR, "Makevars")
if (!file.exists(M)) file.create(M)
arch <- ifelse(R.version$arch == "aarch64", "arm64", "x86_64")
cat(paste("\nCXX14FLAGS += -O3 -mtune=native -arch", arch, "-ftemplate-depth-256"),
    file = M, sep = "\n", append = FALSE)
```

* To install `rstan`, in the R console
  a. Remove all old `rstan` files, if they exist

```{r}
#| eval: false
remove.packages("rstan")
if (file.exists(".RData")) file.remove(".RData")
```

  b. Restart R
  c. Install `StanHeaders` and `rstan`
  
```{r}
#| eval: false
install.packages("StanHeaders", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
install.packages("rstan", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
```

* Verify installation ([details](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started))

```{r}
#| eval: false
remove.packages("rstan")
if (file.exists(".RData")) file.remove(".RData")

install.packages("StanHeaders", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
install.packages("rstan", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
```

* Install brms ([details](https://learnb4ss.github.io/learnB4SS/articles/install-brms.html))

```{r}
#| eval: false
install.packages("brms")
```

You're all set!


## For Windows Users

* Make sure R is up-to-date > 4.2
* [Configural C++ Toolchain](https://github.com/stan-dev/rstan/wiki/Configuring-C---Toolchain-for-Windows) (follow steps for R 4.2)
  * Install RTools42 ([details](https://cran.r-project.org/bin/windows/Rtools/rtools42/rtools.html); [installer](https://cran.r-project.org/bin/windows/Rtools/rtools42/files/rtools42-5355-5357.exe))
* To install `rstan`, in the R console
  a. Remove all old `rstan` files, if they exist

```{r}
#| eval: false
remove.packages("rstan")
if (file.exists(".RData")) file.remove(".RData")
```

  b. Restart R
  c. Install `StanHeaders` and `rstan`
  
```{r}
#| eval: false
install.packages("StanHeaders", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
install.packages("rstan", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
```

* Verify installation ([details](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started))

```{r}
#| eval: false
remove.packages("rstan")
if (file.exists(".RData")) file.remove(".RData")

install.packages("StanHeaders", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
install.packages("rstan", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
```

* Install brms ([details](https://learnb4ss.github.io/learnB4SS/articles/install-brms.html))

```{r}
#| eval: false
install.packages("brms")
```

You're all set!

